Supplement: Impact of seasonal forcing on a potential SARS-CoV-2 pandemic

### 2020-03-11 manuscript/main.tex
resub

\begin{abstract}
A novel coronavirus (SARS-CoV-2) first detected in Wuhan, China, has spread rapidly since December 2019, causing more than 80,000 confirmed infections and 2,700 fatalities (as of Feb 27, 2020). Imported cases and transmission clusters of various sizes have been reported globally suggesting a pandemic is likely.

Here, we explore how seasonal variation in transmissibility could modulate a SARS-CoV-2 pandemic.
Data from routine diagnostics show a strong and consistent seasonal variation of the four endemic coronaviruses (229E, HKU1, NL63, OC43) and we parameterize our model for SARS-CoV-2 using these data.
The model allows for many subpopulations of different size with variable parameters.
Simulations of different scenarios show that plausible parameters result in a small peak in early 2020 in temperate regions of the Northern Hemisphere and a larger peak in winter 2020/2021. 
A smaller range of parameters suggests a dominant peak in the dominant first half of 2020 or two peaks of similar magnitude. 
Variation in transmission and migration rates can result in substantial variation in prevalence between regions.

@@

On Jan 30, 2020, the World Health Organisation (WHO) declared the spread of a new coronavirus, SARS-CoV-2 \citep{ICTV_SARS-CoV-2}, as a public health emergency of international concern \citep{WHO_statement}.
The virus was first identified in patients with pneumonia in the city of Wuhan in the Hubei province, China, in December 2019 \citep{Liangjun2020rna}.
The clinical presentation of the illness caused by SARS-CoV-2, called COVID-19, appears to range from mild or asymptomatic to severe and fatal respiratory illness \citep{WHO_situation_report23}, but the exact spectrum of disease presentation is still unclear. 
The potential for global spread, i.e.~a pandemic, of SARS-CoV-2 is currently not known, but the virus has spread at an alarming rate in Wuhan, the epicenter of the outbreak. Furthermore, the virus has spread to all provinces of China and small clusters of local spread have been reported from several countries, e.g. Singapore, Germany, and the UK \citep{WHO_situation_report23, rothe2020transmission, MOH_Singapore}.

@@

The fraction of tests that were positive for the four seasonal CoVs showed a strong and consistent seasonal variation, see Fig.~\ref{fig:seasonal_CoV}. 
From December to April approximately 2\%  of tests were positive, while less than 0.2\% of tests were positive between July to September, i.e.~a 10-fold difference (Fig.~\ref{fig:seasonal_CoV}, right).
The strength of variation of the transmission rate through the year could be of high relevance to the spread of SARS-CoV-2 in 2020 and following years.

@@

\section{Basic model}
We consider simple SIR models \citep{kermack_contributions_1991} with an additional category $E$ of exposed individuals of the form 
\begin{equation}
\begin{split}
\frac{d}{dt} S & = b(1-S) - \beta(t) S I \\
\frac{d}{dt} E & = \beta(t) S I - \mu E\\
\frac{d}{dt} I & = \mu E - (\nu+b) I \\
R & = 1-S-I-E
\end{split}
\label{eq:SIR}
\end{equation}
where $\beta(t)$ is the rate at which an infected individual infects a susceptible one, $\mu$ is the inverse incubation time, $\nu$ is the recovery rate and $b$ is the population turn-over rate. 
Depending on the analysis below, we implement several such populations that exchange individuals through migration, for details see Supplementary Methods. 
Stochasticity is implemented through Poisson resampling of the population once every serial interval $\mu^{-1}+(\nu+b)^{-1}$.
The population turnover $b$ rate is immaterial for a pandemic scenario, but important for our analysis of seasonal CoVs, and should be interpreted as the sum of the birth rate and the rate at which previously immune individuals become susceptible due to immune waning and escape.
We review general properties of such model in the Supplementary Materials. 
Following previous work, we parameterize transmissibility as 
\begin{equation}
\beta(t) = \beta_0\left(1+\forcing\cos(2\pi (t-\theta))\right)
\label{eq:transmissibility}
\end{equation}
where $\beta_0$ is the average annual infection rate, $\forcing$ is the amplitude of seasonal forcing which modulates transmissibility through the year, and $\theta$ is the time of peak transmissibility \citep{dushoff_dynamical_2004,chen_regular_2017}.
For simulations of the pandemic, we will add an additional term to $\beta(t)$ that accounts for infection control measures in heavily affected areas, see Supplementary Materials.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/fit_2.3.pdf}
    \caption{{\bf Compatibility of SIER model trajectories with observations.} The heatmap shows the inverse mean squared deviation between the model trajectories and the observed seasonal forcing in seasonal CoV prevalence. Model and observations are compatible (yellow shading) in a region of parameter values corresponding to low migration/weak seasonality and second region at high migration/strong seasonality. Migration refers to the rate per year of a susceptible to return from a abroad with a CoV infection.}
    \label{fig:parameter_seasonal}
\end{figure}

@@

These assumptions are not implausible but not certain: cases of SARS-CoV-2 has been in several countries in Asia apart from China, as well as in Europe, Africa, North America, and Australia \citep{WHO_situation_report37}, and mild or asymptomatic cases make detection and thus prevention of spread by isolation challenging, \textit{e.g.} airport screening as a preventive measure is unlikely to prevent spread and local seeding \citep{quilty2020effectiveness}. 
Person-to-person transmission of the virus has been documented in several countries outside of China, including large outbreaks in Iran, South Korea, and Italy \citep{ECDC_23feb}. 
It is likely that not all exported cases have been detected, and some may have seeded outbreaks outside of China that have yet to be detected.

>>>>

\begin{abstract}
A novel coronavirus (SARS-CoV-2) first detected in Wuhan, China, has spread rapidly since December 2019, causing more than 100,000 confirmed infections and 4,000 fatalities (as of Mar 10, 2020). Imported cases and transmission clusters of various sizes have been reported globally suggesting a pandemic is likely.

Here, we explore how seasonal variation in transmissibility could modulate a SARS-CoV-2 pandemic.
Data from routine diagnostics show a strong and consistent seasonal variation of the four endemic coronaviruses (229E, HKU1, NL63, OC43) and we parameterize our model for SARS-CoV-2 using these data.
The model allows for many subpopulations of different size with variable parameters.
Simulations of different scenarios show that plausible parameters result in a small peak in early 2020 in temperate regions of the Northern Hemisphere and a larger peak in winter 2020/2021. 
Variation in transmission and migration rates can result in substantial variation in prevalence between regions.

@@

On Jan 30, 2020, the World Health Organisation (WHO) declared the spread of a new coronavirus, SARS-CoV-2 \citep{ICTV_SARS-CoV-2}, as a public health emergency of international concern \citep{WHO_statement}, and on March 11 as a pandemic \citep{WHOpandemic}.
The virus was first identified in patients with pneumonia in the city of Wuhan in the Hubei province, China, in December 2019 \citep{Liangjun2020rna}.
The clinical presentation of the illness caused by SARS-CoV-2, called COVID-19, appears to range from mild or asymptomatic to severe and fatal respiratory illness \citep{WHO_China2020joint}, but the exact spectrum of disease presentation is still unclear. 
The potential for global spread, i.e.~a pandemic, of SARS-CoV-2 is currently not known, but the virus has spread at an alarming rate in Wuhan, the epicenter of the outbreak. Furthermore, the virus has spread to all provinces of China and small clusters of local spread have been reported from several countries, e.g. Singapore, Germany, the UK, and Italy \citep{WHO_situation_report23, rothe2020transmission, MOH_Singapore, ECDC_report9March}.

@@

The fraction of tests that were positive for the four seasonal CoVs showed a strong and consistent seasonal variation, see Fig.~\ref{fig:seasonal_CoV}. 
From December to April approximately 2\%  of tests were positive, while less than 0.2\% of tests were positive between July to September, i.e.~a 10-fold difference (Fig.~\ref{fig:seasonal_CoV}, right).
While the fraction of positive tests is not a direct measure of prevalence, it is a reasonable proxy that we expect to be well correlated with prevalence. 
The strength of variation of the transmission rate through the year could be of high relevance to the spread of SARS-CoV-2 in 2020 and following years.

@@

\section{Basic model}
We consider simple SIR models \citep{kermack_contributions_1991} with an additional category $E$ of exposed individuals of the form 
\begin{equation}
\begin{split}
\frac{d}{dt} S & = b(1-S) - \beta(t) S I \\
\frac{d}{dt} E & = \beta(t) S I - (\mu +b) E\\
\frac{d}{dt} I & = \mu E - (\nu+b) I \\
R & = 1-S-I-E
\end{split}
\label{eq:SIR}
\end{equation}
where $\beta(t)$ is the rate at which an infected individual infects a susceptible one, $\mu$ is the inverse incubation time, $\nu$ is the recovery rate and $b$ is the population turn-over rate. 
Depending on the analysis below, we implement several such populations that exchange individuals through migration, for details see Supplementary Methods. 
Stochasticity is implemented through Poisson resampling of the population once every serial interval $\mu^{-1}+(\nu+b)^{-1}$.
The population turnover $b$ rate is immaterial for a pandemic scenario, but important for our analysis of seasonal CoVs, and should be interpreted as the sum of the birth rate and the rate at which previously immune individuals become susceptible due to immune waning and escape.
We review general properties of such model in the Supplementary Materials. 
Following previous work, we parameterize transmissibility as 
\begin{equation}
\beta(t) = \beta_0\left(1+\forcing\cos(2\pi (t-\theta))\right)
\label{eq:transmissibility}
\end{equation}
where $\beta_0$ is the average annual infection rate, $\forcing$ is the amplitude of seasonal forcing which modulates transmissibility through the year, and $\theta$ is the time of peak transmissibility \citep{dushoff_dynamical_2004,chen_regular_2017}.
For simulations of the pandemic, we will add an additional term to $\beta(t)$ that accounts for infection control measures in heavily affected areas, see Supplementary Materials.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/fit_2.3.pdf}
    \caption{{\bf Compatibility of SEIR model trajectories with observations.} The heatmap shows the inverse mean squared deviation between the model trajectories and the observed seasonal forcing in seasonal CoV prevalence. Model and observations are compatible (yellow shading) in a region of parameter values corresponding to low migration/weak seasonality and second region at high migration/strong seasonality. Migration refers to the rate per year of a susceptible to return from a abroad with a CoV infection.}
    \label{fig:parameter_seasonal}
\end{figure}

@@

These assumptions are not implausible but not certain: cases of SARS-CoV-2 has been in several countries in Asia apart from China, as well as in Europe, Africa, North America, and Australia \citep{WHO_situation_report50}, and mild or asymptomatic cases make detection and thus prevention of spread by isolation challenging, \textit{e.g.} airport screening as a preventive measure is unlikely to prevent spread and local seeding \citep{quilty2020effectiveness}. 
Person-to-person transmission of the virus has been documented in several countries outside of China, including large outbreaks in Iran, South Korea, and Italy \citep{WHO_situation_report50}. 
It is likely that not all exported cases have been detected, and some may have seeded outbreaks outside of China that have yet to be detected.

### 2020-03-11 manuscript/supplement.tex
resub

\setcounter{figure}{0}
\renewcommand{\figurename}{Figure S}
\setcounter{table}{0}
\renewcommand{\tablename}{Table S}
\maketitle
The SIR model introduced in equations \ref{eq:SIR} describes the fraction of susceptible $S(t)$, exposed $E(t)$, and infected individuals $I(t)$ in a population. 
The fraction of recovered individuals $R$ is $R=1 - S - E - I$.

An illustration of the model is presented in figure \ref{fig:schemeSIR} with constant transmissibility ($\forcing=0$ in equation \ref{eq:transmissibility}).

@@

\begin{figure}[htb]
	\centering
	\includegraphics[width=0.9\columnwidth]{figures/SEIR_sketch.pdf}
	\caption{Illustration of the SIER model used with its parameters.}\label{fig:schemeSIR}
\end{figure}

For our analysis of seasonal CoVs, it is instructive to study the stability of this steady state and show that the quantities $[S,I]$ tend to show damped oscillations around the fixed points in Eq.~\ref{eq:fixed_point} with a period
\begin{equation}
T = \frac{4\pi}{\sqrt{4b(\nu(R_0-1) - b) - R_0^2b^2}}
\label{eq:period}
\end{equation}
This behavior is intrinsic to the system and holds whenever transmissibility is constant.

@@

In a fully susceptible population, the number of infected individuals initially grows as $I(t)\sim e^{(\beta-\nu-b)t} = e^{(R_0-1)t/(\nu+b)}$.
The doubling time is $\tau_2 = (\nu+b)\log(2)/(R_0-1)$.
Since $b\ll \nu$, we can safely neglect $b$ in the early phase of the pandemic, but it is important when interpreting endemic seasonal CoV data.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/fraction_positive_HKU1_OC43.pdf}
    \caption{Seasonal variation of HKU1/OC43 CoV positive tests in Stockholm, Sweden. }
    \label{fig:seasonal_variation_supp}
\end{figure}

>>>>

\setcounter{figure}{0}
\renewcommand{\figurename}{Figure S}
\setcounter{table}{0}
\renewcommand{\tablename}{Table S}
\maketitle
The SIR model introduced in equations \ref{eq:SIR} describes the fraction of susceptible $S(t)$, exposed $E(t)$, and infected individuals $I(t)$ in a population. 
The fraction of recovered individuals $R$ is $R=1 - S - E - I$. 
An illustration of the model is presented in figure \ref{fig:schemeSIR}.

\begin{figure}[htb]
	\centering
	\includegraphics[width=0.9\columnwidth]{figures/SEIR_sketch.pdf}
	\caption{Illustration of the SEIR model used with its parameters.}\label{fig:schemeSIR}
\end{figure}

@@

For our analysis of seasonal CoVs, it is instructive to study the stability of this steady state of the simplified SIR (not SEIR) model. 
It is easy to show that the quantities $[S,I]$ tend to show damped oscillations around the fixed points in Eq.~\ref{eq:fixed_point} with a period
\begin{equation}
T = \frac{4\pi}{\sqrt{4b(\nu(R_0-1) - b) - R_0^2b^2}}
\label{eq:period}
\end{equation}
This behavior is intrinsic to the system and holds whenever transmissibility is constant.

@@

Similarly, in the simplified SIR (not SEIR) model, the initial growth of the number of cases can be expressed analytically.
In a fully susceptible population, the number of infected individuals initially grows as $I(t)\sim e^{(\beta-\nu-b)t} = e^{(R_0-1)t/(\nu+b)}$.
The doubling time is $\tau_2 = (\nu+b)\log(2)/(R_0-1)$.
Since $b\ll \nu$, we can safely neglect $b$ in the early phase of the pandemic, but it is important when interpreting endemic seasonal CoV data. 
For detailed discussion of the effect of generation times and incubation periods on the initial exponential growth, see \citep{wallinga_how_2007}.

@@

\bibliography{bib}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/fraction_positive_HKU1_OC43.pdf}
    \caption{Seasonal variation of HKU1/OC43 CoV positive tests in Stockholm, Sweden. }
    \label{fig:seasonal_variation_supp}
\end{figure}
